---
title: "Survey demographics"
output: 
  html_document: 
    df_print: default
    toc: yes
  html_notebook: 
    toc: yes
---

<center> *David Doret* </center>
<center> *Version 1.0* </center>
<center> *June 2020* </center>

## In a nutshell

In this R notebook, ...

## Introduction

As part of the *IAM Performance Measurement 2020 Survey*, I surveyed IAM professionals ...

This notebook focuses on data analysis. Its business implications are discussed in a distinct article.

## Setting up the technical environment

To conduct our data analysis, we first need to setup our R technical environment.

```{r echo=TRUE, message=FALSE}

# Set console to English
Sys.setenv(LANG = "en");

# Install R packages as needed
if(!require("devtools")) install.packages("devtools");
#if(!require("eulerr")) install.packages("eulerr");
#if(!require("tidyr")) install.packages("tidyr");
#if(!require("RCurl")) install.packages("RCurl");
#if(!require("plyr")) install.packages("plyr");
#if(!require("dplyr")) install.packages("dplyr");
#if(!require("knitr")) install.packages("knitr");
#if(!require("stringr")) install.packages("stringr");
#if(!require("remotes")) install.packages("remotes");
#if(!require("stringr.tools")) #remotes::install_github("decisionpatterns/stringr.tools");
#if(!require("sjlabelled")) install.packages("sjlabelled");
#if(!require("naniar")) install.packages("naniar");
#if(!require("tidyverse")) install.packages("tidyverse");
#if(!require("likert")) install.packages("likert");
#if(!require("matrixStats")) install.packages("matrixStats");
#if(!require("sjmisc")) install.packages("sjmisc");
#if(!require("ggplot2")) install.packages("ggplot2");
#if(!require("ggpubr")) install.packages("ggpubr");
#if(!require("VGAM")) install.packages("VGAM");
#if(!require("gridExtra")) install.packages("gridExtra");

```

## Loading the IAMPerf2020 data environment

```{r}
devtools::source_url("https://raw.githubusercontent.com/Open-Measure/Open-Data/master/IAMPerf2020-Dataset/IAMPerf2020Setup.R?raw=TRUE")
```

# Demographics



# Answers

## Devices

```{r}

get_device_plot_data = function(survey_data){
  incomplete = table(survey_data[survey_data$ResponseStatus != "Complete",]$Device);
  complete = table(survey_data[survey_data$ResponseStatus == "Complete",]$Device);
  
  plot_data = data.frame(
    fill = c(names(incomplete), names(complete)),
    y = c(incomplete, complete),
    x = c(rep("Incomplete", 3), rep("Complete", 3))
  );
  
  label_percent = c(
    paste(rounded_ratios_with_largest_remainder(incomplete, 100, 2), "%", sep = ""),
    paste(rounded_ratios_with_largest_remainder(complete, 100, 2), "%", sep = "")
    );
  label_absolute = paste("(", plot_data$y, ")", sep = "");
  label = paste(label_percent, label_absolute, sep = "\n");
  plot_data$label = label;

  return(plot_data);  
};

# Stacked + percent
ggplot2::ggplot(
  get_device_plot_data(iamperf2020_survey), 
  ggplot2::aes(fill=fill, y=y, x= x, label = label)
  ) + 
  ggplot2::geom_bar(position="fill", stat="identity") +
  ggplot2::geom_text(size = 3, position = ggplot2::position_fill(vjust = 0.5)) + 
  ggplot2::xlab("Survey status") +
  ggplot2::scale_x_discrete (limits = c("Incomplete", "Complete")) +
  ggplot2::scale_y_continuous (name = "%") +
  ggplot2::scale_fill_brewer(palette = "Paired") +
  ggplot2::labs(title = "Participant devices", subtitle = "The device category used by survey participants", fill = "Device")

```

# Survey Duration

```{r}

light_grey = "#dddddd";
light_blue = "#5599ff";
dark_blue = "#0066ff";

get_timetaken_plot_data = function(survey_data){
  
  time_taken = data.frame(
    x = iamperf2020_survey$ResponseStatus,
    y = iamperf2020_survey$TimeTaken / 60,
    fill = iamperf2020_survey$ResponseStatus
  );

};


ggplot2::ggplot(
  get_timetaken_plot_data(iamperf2020_survey), 
  ggplot2::aes(x = x, y = y, fill = fill)) +
  ggplot2::scale_fill_brewer(palette = "Paired") +
  ggplot2::stat_boxplot(
    width = 0.20,
    size = 1.25,
    geom = "errorbar") + 
  ggplot2::geom_boxplot(
    outlier.shape = NA,
    lwd = 0.2,
    fill = "#FFFFFF") + 
  ggplot2::geom_jitter(
    size = 3,
    shape = 21,
    #fill = light_blue,
    colour = "#000000") +
 ggplot2::xlab(label = "Survey status") +
 ggplot2::ylab(label = "Survey duration (in minutes)") +
 ggplot2::coord_flip() +
 ggplot2::labs(title = "Survey Duration", subtitle = "The distribution of survey durations", fill = "Survey status")


```



# References

I list here a bunch of articles that have been inspiring or helpful while writing this notebook.

 * https://stackoverflow.com/questions/35720660/how-to-use-an-r-script-from-github
 