---
title: "Survey demographics"
output: 
  html_document: 
    df_print: default
    toc: yes
  html_notebook: 
    toc: yes
---

<center> *David Doret* </center>
<center> *Version 1.0* </center>
<center> *June 2020* </center>

## In a nutshell

In this R notebook, ...

## Introduction

As part of the *IAM Performance Measurement 2020 Survey*, I surveyed IAM professionals ...

This notebook focuses on data analysis. Its business implications are discussed in a distinct article.

## Setting up the technical environment

To conduct our data analysis, we first need to setup our R technical environment.

```{r echo=TRUE, message=FALSE}

# Set console to English
Sys.setenv(LANG = "en");

# Install R packages as needed
if(!require("devtools")) install.packages("devtools");
#if(!require("eulerr")) install.packages("eulerr");
#if(!require("tidyr")) install.packages("tidyr");
#if(!require("RCurl")) install.packages("RCurl");
#if(!require("plyr")) install.packages("plyr");
#if(!require("dplyr")) install.packages("dplyr");
#if(!require("knitr")) install.packages("knitr");
#if(!require("stringr")) install.packages("stringr");
#if(!require("remotes")) install.packages("remotes");
#if(!require("stringr.tools")) #remotes::install_github("decisionpatterns/stringr.tools");
#if(!require("sjlabelled")) install.packages("sjlabelled");
#if(!require("naniar")) install.packages("naniar");
#if(!require("tidyverse")) install.packages("tidyverse");
#if(!require("likert")) install.packages("likert");
#if(!require("matrixStats")) install.packages("matrixStats");
#if(!require("sjmisc")) install.packages("sjmisc");
#if(!require("ggplot2")) install.packages("ggplot2");
#if(!require("ggpubr")) install.packages("ggpubr");
#if(!require("VGAM")) install.packages("VGAM");
#if(!require("gridExtra")) install.packages("gridExtra");

```

Some will notice I don't load libraries. Instead, I prefer to use the unambiguous syntax *package::function*. This makes the code slightly harsher to read but this is a price I am pleased to pay. 

## Loading complementary scripts from Github
# devtools::source_url("https://github.com/tonybreyal/Blog-Reference-Functions/blob/master/R/bingSearchXScraper/bingSearchXScraper.R?raw=TRUE")

## Retrieving the data

Then we need to load our survey data. In accordance with the *Open Source* and *Open Data* spirit of the *Open-Measure* project, all data is publicly available on GitHub. Isn't that extremely cool, hm?

```{r}





```


# References

I list here a bunch of articles that have been inspiring or helpful while writing this notebook.

 * https://stackoverflow.com/questions/35720660/how-to-use-an-r-script-from-github